---
title: "Workshop 1 write-up"
author: "Jorge Valdes"
date: "9/26/2018"
output:
  word_document: default
  html_document: default
---

```{r import data}
library(tidyverse)
cs_reading <- read.csv("L2CS_Word8_Ps11_20.csv")
str(cs_reading)
```
```{r}
names(cs_reading)
levels(cs_reading$sentence_type)
```

We have to identify the following columns for our new reduced dataset:
* RECORDING_SESSION_LABEL
* aux_type
* switch_type
* sentence_type
* button_RT
* response_accuracy
* GAZE_DURATION
* REGRESSION_PATH_DURATION
* TOTAL_DURATION

We will use the `select()` command to narrow our dataset. This will be saved to a new dataset. 

```{r select dataset}
cs_reading2 <- cs_reading %>% 
  select(RECORDING_SESSION_LABEL, aux_type, switch_type, sentence_type, button_RT, response_accuracy, GAZE_DURATION, REGRESSION_PATH_DURATION, TOTAL_DURATION)
head(cs_reading2)
```

We noticed that for Windows machines, there was an issue with how the column header for the first column was encoded. This involved a weird character. One way to check this is to use the `names()` command which lists just the column names (check the first column header name for any weird character). 

```{r}
names(cs_reading)
```

For Windows users, one quick solution is to manually rewrite the first column header. We can do this with the following code (the [1] means that we are rewriting the name of the first column header). This should then let one proceed with code as usual. 

```{r}
names(cs_reading)[1] <- "RECORDING_SESSION_LABEL"
head(cs_reading)
```

We only want experimental sentences. We will use `filter()` to select just experimental sentences.

```{r filter}
cs_reading3 <- cs_reading2 %>% 
  filter(sentence_type == "Experimental" & response_accuracy == "correct")
head(cs_reading3)
```

Now we are going to summarize the data by the variables of interest, aux_type and switch_type. 

```{r summarize}
cs_summary <- cs_reading3 %>% 
  group_by(aux_type, switch_type) %>% 
  summarize(mGAZE_DURATION = mean(GAZE_DURATION), mREGRESSION_PATH_DURATION = mean(REGRESSION_PATH_DURATION), mTOTAL_DURATION = mean(TOTAL_DURATION), sdGAZE_DURATION = sd(GAZE_DURATION), sdREGRESSION_PATH_DURATION = sd(REGRESSION_PATH_DURATION), sdTOTAL_DURATION = sd(TOTAL_DURATION))
cs_summary

```

We also collected button RT from the amount of time it took to answer the comprehension question. Now we will summarize and get the mean and sd of button_RT. We noticed in class that for some of us, the button_RT was not summarizing with mean. It turns out that the variable was considered a factor (instead of an integer), so we first had to change the class to an integer class using `as.integer()`. 

```{r}
cs_reading3$button_RT <- as.integer(cs_reading3$button_RT)

buttonrt_summary <- cs_reading3 %>% 
  group_by(aux_type, switch_type) %>% 
  summarize(mbuttonRT = mean(button_RT), sdbuttonRT = sd(button_RT))
buttonrt_summary
```

let's graph mean button press RTs very quickly.

```{r}
ggplot(buttonrt_summary, aes(aux_type, mbuttonRT, fill = switch_type)) +
  geom_bar(stat = "identity", position = "dodge")
```

